Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Clique Neighbourhood Constraints

### 2021-01-20 paper/paper.tex
Completely anonymous
Tweaks

% \usepackage[textsize=tiny,textwidth=2cm,shadow]{todonotes}
% \newcommand{\skcom}[1]{\todo[color=orange!50!white]{{S}: #1}}
% \newcommand{\rcom}[1]{\todo[color=gray!50!white]{{R}: #1}}%reviewers said
% \usepackage{showframe}

@@

% \author{
% Ciaran McCreesh$^1$
% \and
% Sonja Kraiczy$^2$
% \affiliations
% $^1$University of Glasgow\\
% $^2$University of Oxford\\
% \emails
% ciaran.mccreesh@glasgow.ac.uk
% }

@@

\begin{abstract}
    Graph homomorphism problems involve finding adjacency-preserving mappings between two graphs.
    Many of these problems are theoretically hard, but can be solved quite efficiently in practice
    through constraint programming techniques.  This paper adapts techniques from the
    state-of-the-art in subgraph isomorphism solving to broader graph homomorphism problems, and
    introduces a new form of filtering based upon clique-finding. We evaluate these techniques
    empirically, and show they are effective in practice for the locally injective graph
    homomorphism and subgraph isomorphism problems, as well as giving the first practical constraint
    programming approach to finding general graph homomorphisms.
\end{abstract}

@@

The \emph{subgraph isomorphism problem}, as defined by \citet{DBLP:books/fm/GareyJ79}, is to
determine whether an injective mapping exists from one given graph to another, such that adjacent
pairs of vertices are mapped to adjacent pairs of vertices, and non-adjacent vertices are mapped to
non-adjacent vertices.  However, in application-oriented papers, particularly from bioinformatics
and chemistry, the same name is often implicitly used to mean the \emph{non-induced} version of the
problem (which does not require that non-adjacency be preserved) since this more accurately models
the real-world problem being solved \cite{Willett1999,EhrlichR11}. Despite being theoretically hard
problems, these applications, along with
others in areas including compilers \cite{DBLP:conf/cp/BlindellLCS15}
and pattern recognition \cite{DBLP:journals/ijprai/FoggiaPV14},
have given rise to a large amount of research into designing practical algorithms for solving these
problems.  Most approaches are based either upon very fast but simple backtracking algorithms
\cite{DBLP:journals/pami/CordellaFSV04,DBLP:journals/bmcbi/BonniciGPSF13,DBLP:conf/gbrpr/CarlettiFSV17}
which often but not always perform well on very easy instances, or upon constraint programming
algorithms
\cite{DBLP:journals/constraints/ZampelliDS10,DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/cpaior/ArchibaldDHMP019},
which have higher startup costs but that perform vastly better on harder instances and much more
consistently on easy instances \cite{DBLP:journals/jair/McCreeshPST18,DBLP:conf/gbrpr/Solnon19}.

@@

These injectivity relaxations have received much less attention
from a practical solving perspective.  One might hope that the inference techniques developed for
subgraph isomorphism would also be helpful for other graph homomorphism problem variants. Indeed,
this paper proves that many, but not all, of these strategies are also valid in the locally
injective case, and that simple distance filtering is valid for all homomorphisms.  However, we also
show that \emph{none} of the other strategies are valid for finding homomorphisms where there is no
injectivity requirement.  Finally, we introduce a new filtering technique that is based upon finding
a maximum clique in the neighbourhood of each domain vertex, which is valid even in the general
case. Although this new filtering technique involves solving many additional \NP-complete problems
as a preprocessing step, we demonstrate that it is effective in practice, particularly for the
non-injective problem where filtering allows a constraint programming algorithm to solve instances
over eight hundred times faster in aggregate, over a collection of over fourteen thousand standard
benchmark instances. This shows, for the first time, that modern constraint programming techniques
can be practical for less constrained graph homomorphism finding problems; previous algorithmic
approaches have focused instead upon worst-case computational bounds
(e.g.\
\citet{DBLP:journals/mst/FominHK07,DBLP:journals/csr/FialaK08,DBLP:journals/ipl/Rzazewski14,DBLP:journals/tcs/ChaplickFHPT15}),
whose practical utility has yet to be demonstrated. The empirical effectiveness of our results is
especially important because many current applications use subgraph isomorphism solvers only because
they perform well using off the shelf solvers, rather than because they exactly match ideal domain
requirements.

@@

\begin{proof}
    Let $j$ and $k$ be two distinct vertices of the clique in $G$. By definition of a homomorphism,
    $i(j)$ and $i(k)$ must be adjacent. And, because $H$ does not contain loops, $i(j)$ cannot equal
    $i(k)$.
\end{proof}

@@

We then move on to the codomain vertices, using a similar caching routine. Rather than calculating a
clique size for every single codomain vertex, we only calculate a value for codomain vertices which are
present in at least one variable's domain, after other unary constraints have been applied.
Additionally, we do not require the maximum clique solver to run to completion and guarantee that it
has found a maximum clique. Instead, we allow it to stop as soon as it has found a clique with as
many vertices as the largest domain clique. This is useful in practice because it may be very hard
to decide whether a particular codomain vertex has neighbourhood clique size of, say, 15 or 16, but if
the largest domain vertex has a neighbourhood clique size of only 5 then this is irrelevant.

@@

To quantify these gains, the top centre plot of \cref{figure:plots} shows the
additional number of instances solved at any given time (that is, the vertical distance between
two cumulative plots) when enabling additional filtering. We see that for subgraph isomorphism, at
the one hour timeout we solve 22 additional instances using clique filtering; for locally injective
homomorphisms, 21 additional instances; and for homomorphisms, 114, 146 or 193 additional instances
using just distance filtering, just clique filtering, or both. Alternatively, we can measure the
horizontal distance between plots for a given choice of timeout $t$ as follows. Let $t'$
be the last time no greater than $t$ where the slower algorithm solved an instance, and let $u$ be
the lowest amount of time needed for the faster algorithm to solve the same number of instances.
The \emph{aggregate speedup}, which we plot in the top right of \cref{figure:plots}, is the ratio of
$t'$ to $u$ or $u$ to $t'$ as appropriate. At the one hour timeout, we see aggregate speedups of 2.2
for locally injective homomorphisms, 3.2 for subgraph isomorphism, and 47.5, 131.7, and 803.5 for
homomorphisms.

@@

% \section*{Acknowledgements}
% 
% This work was supported by the Engineering and Physical Sciences Research Council [grant number
% EP/P026842/1].

>>>>

@@

@@

\begin{abstract}
    Graph homomorphism problems involve finding adjacency-preserving mappings between two given
    graphs.  Although theoretically hard, these problems can often be solved in practice
    using constraint programming algorithms. We show how techniques from the state-of-the-art in
    subgraph isomorphism solving can be applied to broader graph homomorphism problems, and
    introduce a new form of filtering based upon clique-finding. We demonstrate empirically that
    this filtering is effective for the locally injective graph homomorphism and subgraph
    isomorphism problems, and gives the first practical constraint programming approach to
    finding general graph homomorphisms.
\end{abstract}

@@

The \emph{subgraph isomorphism problem}, as defined by \citet{DBLP:books/fm/GareyJ79}, is to
determine whether an injective mapping exists from one given graph to another, such that adjacent
pairs of vertices are mapped to adjacent pairs of vertices, and non-adjacent vertices are mapped to
non-adjacent vertices.  However, in application-oriented papers, particularly from bioinformatics
and chemistry, the same name is often implicitly used to mean the \emph{non-induced} version of the
problem (which does not require that non-adjacency be preserved) since this more accurately models
the real-world problem being solved \cite{Willett1999,EhrlichR11}. Despite being theoretically hard
problems, these applications, along with
others in areas including compilers \cite{DBLP:conf/cp/BlindellLCS15},
graph databases \cite{DBLP:journals/jair/McCreeshPST18}
and pattern recognition \cite{DBLP:journals/ijprai/FoggiaPV14},
have given rise to a large amount of research into designing practical algorithms for solving these
problems.  Most approaches are based either upon very fast but simple backtracking algorithms
\cite{DBLP:journals/pami/CordellaFSV04,DBLP:journals/bmcbi/BonniciGPSF13,DBLP:conf/gbrpr/CarlettiFSV17}
which often but not always perform well on very easy instances, or upon constraint programming
algorithms
\cite{DBLP:journals/constraints/ZampelliDS10,DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/cpaior/ArchibaldDHMP019},
which have higher startup costs but that perform vastly better on harder instances and much more
consistently on easy instances \cite{DBLP:journals/jair/McCreeshPST18,DBLP:conf/gbrpr/Solnon19}.

@@

These injectivity relaxations have received much less attention
from a practical solving perspective.  One might hope that the inference techniques developed for
subgraph isomorphism would also be helpful for other graph homomorphism problem variants. Indeed,
this paper proves that many, but not all, of these strategies are also valid in the locally
injective case, and that simple distance filtering is valid for all homomorphisms.  However, we also
show that \emph{none} of the other strategies are valid for finding homomorphisms where there is no
injectivity requirement.  Finally, we introduce a new filtering technique that is based upon finding
a maximum clique in the neighbourhood of each domain vertex, which is valid even in the general
case. Although this new filtering technique involves solving many additional \NP-complete problems
as a preprocessing step, we demonstrate that it is effective in practice, particularly for the
non-injective problem where filtering allows a constraint programming algorithm to solve instances
over eight hundred times faster in aggregate, over a collection of over fourteen thousand standard
benchmark instances. This shows, for the first time, that modern constraint programming techniques
can be practical for less constrained graph homomorphism finding problems; previous algorithmic
approaches have focused instead upon worst-case computational bounds
(e.g.\
\citet{DBLP:journals/mst/FominHK07,DBLP:journals/csr/FialaK08,DBLP:journals/ipl/Rzazewski14,DBLP:journals/tcs/ChaplickFHPT15}),
whose practical utility has yet to be demonstrated. The empirical effectiveness of our results is
especially important because many applications use subgraph isomorphism solvers only because
they perform well off the shelf, rather than because they exactly match domain requirements.

@@

\begin{proof}
    Let $j$ and $k$ be two distinct vertices of the clique in $G$. By definition of a homomorphism,
    $i(j)$ and $i(k)$ must be adjacent. And, because $H$ has no loops, $i(j) \ne i(k)$.
\end{proof}

@@

We then move on to the codomain vertices, using a similar caching routine. Rather than calculating a
clique size for every single codomain vertex, we only calculate a value for codomain vertices which are
present in at least one variable's domain, after other unary constraints have been applied.
Additionally, we do not require the maximum clique solver to run to completion and guarantee that it
has found a maximum clique. Instead, we allow it to stop as soon as it has found a clique with as
many vertices as the largest domain clique. This is useful because it may be very hard to decide
whether a particular codomain vertex has neighbourhood clique size of, say, 15 or 16, but if the
largest domain vertex has a neighbourhood clique size of only 5 then this is irrelevant.

@@

To quantify these gains, the top centre plot of \cref{figure:plots} shows the
additional number of instances solved at any given time (that is, the vertical distance between
two cumulative plots) when enabling additional filtering. We see that for subgraph isomorphism, at
the one hour timeout we solve 22 additional instances using clique filtering; for locally injective
homomorphisms, 21 additional instances; and for homomorphisms, 114, 146 or 193 additional instances
using just distance filtering, just clique filtering, or both. Or, at the best choices of timeouts,
we solve 53 additional subgraph isomorphism instances; 64 additional locally injective instances;
and 243, 383, and 424 additional homomorphism instances.

Alternatively, we can measure the horizontal distance between plots for a given choice of timeout
$t$ as follows. Let $t'$ be the last time no greater than $t$ where the slower algorithm solved an
instance, and let $u$ be the lowest amount of time needed for the faster algorithm to solve the same
number of instances.  The \emph{aggregate speedup}, which we plot in the top right of
\cref{figure:plots}, is the ratio of $t'$ to $u$ or $u$ to $t'$ as appropriate. At the one hour
timeout, we see aggregate speedups of 2.2 for locally injective homomorphisms, 3.2 for subgraph
isomorphism, and 47.5, 131.7, and 803.5 for homomorphisms.

@@
